### 2024-05-21 sciperl/paper.tex
Initial commit

\documentclass{article}
\usepackage{hyperref}
\begin{document}
Howdy \LaTeX

\url{https://troglodyne.net}
\end{document}

### 2024-05-21 sciperl/paper.tex
Rough draft in place for rigor paper.

@@
-\documentclass{article}
+\documentclass[twoside]{article}
 \usepackage{hyperref}
+\usepackage{graphicx}
+\usepackage[dvipsnames]{xcolor}
+\hypersetup{
+     colorlinks=true,
+     allcolors=NavyBlue
+}
+\usepackage{listings}
+\lstset{language=Perl}
+
+\title{%
+    Reasoning about the rigor of perl programs: \\
+    \large gaps, limits to understanding, and means to ameliorate
+}
+\author{%
+    George S. Baugh \\
+    \large CEO, Troglodyne LLC \\
+    \url{https://troglodyne.net}
+}
 \begin{document}
-Howdy \LaTeX
+\maketitle
 
-\url{https://troglodyne.net}
+\section{ABSTRACT}
+
+While some tools exist to reason about the correctness of a program and its fitness for particular purposes,
+there remain a number of unexamined approaches which will likely prove fruitful to explore.
+
+Productivity of any piece of capital or laborer is entirely a function of error rate; being able to quantify
+the conditions under which failures can be expected is thus of the utmost importance.
+
+In the era of testsuite generation via large language models, this is of even greater concern.
+
+I will lay out what is currently available, how that fits into a framework for reasoning about programs
+in a rigorous fashion, and identify places where our ability to reason could be improved via tooling.
+Prototypes of said tools will also be demonstrated.
+
+\section{INTRODUCTION}
+
+In software testing there's a concept known as the "pesticide paradox" which is shorthand for saying
+"The bugs are always where you aren't looking".  Abscence of evidence is not evidence of abscence.
+
+If we can straightforwardly identify where we are not looking, and cross-reference this with
+a means of discerning which bits of code are the most critical we can then prioritize our annealing process.
+
+To identify where we are not looking, we need a proper mental model of how computers work.
+Thankfully, this is well established.  Programs are just stacks of instructions capable of adding and removing instructions to said stack.
+Modern programs are usually multiple series of these stacks in parallel, usually referred to as pipelines.
+Some of these instructions are thought of as code, some as data but it's all the same to the executor.
+
+Any given instruction is going to operate like a traditional mathematical function;
+this is to say that it will have a range of acceptable inputs and outputs.
+Anything outside of these ranges results in either invalid or undefined behavior when composed in a stack of instructions.
+Much like a child's game of "telephone", the earlier in the stack this happens, typically the worse the outcome.
+
+From this two things are clear:
+
+1. rigorous testing of any individual *instruction* within a stack over it's expected domain of inputs, and a sampling of data outside of that is valuable.
+2.
+The interaction of an instruction with its immediate follower and predecessor must also be well understood; these multi-instruction sections can indeed be treated as though they are themselves more complicated instructions.
+
+This is the actual theoretical underpinning behind the two testing concepts we are all familiar with (unit and integration testing), but most don't think too hard about.
+It's this lack of concrete understanding that leads to frequent confusion and ineffective testsuites the world over.
+
+It is tempting from here to consider programs as discrete stacks, which can be exhaustively proven to behave well, in a sort of "proof by induction".
+However the data you feed in is itself a stack of instructions with a necessarily unbounded input and output domain.
+This is best known as the "halting problem", as there is always theoretically some set of data you can feed a program which will prevent any program from ever terminating.
+What is less well understood is that the pesticide paradox is pernicious precisely because it is the halting problem.
+And the halting problem is actually Godel's Uncertainty principle - there will always exist at least one statement in any given axiomatic system which cannot ever be proven.
+
+This same problem rears its head when attempting to demonstrate fitness for a particular purpose!
+Any approach other than actually engaging in said purpose immediately falls afoul of the "frame problem".
+Ultimately nothing but an actual user can actually accept a product is fit for a particular purpose.
+As such, any approach other than recording user journeys and being able to play them back is shooting in the dark.
+
+Economists call this concept "Demonstrated Preference", and the impossibility of centralized economic planning is largely the frame problem restated.
+Many firms fall victim to this conceit with their "user stories" and acceptance tests based around these.
+Ultimately the only means to catch fitness issues early in new features is to eat your own dog food, being particularly attentive to "Schlepping" and other places the work-flow can be de-burred.
+
+\section{Reasoning about Perl Programs}
+
+All programs can after all be expressed as lambda calculus; programs are nearly always transformed into some variety of "intermediate form" prior to being handed off to lower level executors.
+One such form is the "abstract syntax tree", which is an intermediate form used to reason about the correctness of programs rather than to pass to hardware for execution.
+
+Building such a tree is how PPI, and the tools which rely on it such as Perl::Critic work.
+Alternatively you can inspect the "intermediate form" which the interpreter itself builds via the B::Deparse, or similar techniques as Devel::Cover and Devel::NYTProf leverage.
+However just because you *can* reason about perl programs in the same fashion as any other programming language, does not guarantee avenues of inquiry are exhaustively explored.
+
+a. Devel::Cover does not quite cover the subject
+
+Devel::Cover is quite good at identifying statements and conditions which are not covered within a program.
+It cannot however identify places where a condition ought to be applied, but is not.
+
+It also does not have a means of identifying whether the boundaries of any given block's input/output domain have been exercised.
+It is generally via these means that we discover these missed conditionals, as we are frequently surprised by the realities imposed by input data.
+Given perl only has optional input signatures, and no output signatures nor ways to signify that a sub throws, this has to be inferred; a characterization of the stack must be built.
+
+This inference necessarily implies walking the stack down to the most basic of calls.
+Suppose we have a sub foo, which calls bar() that calls baz():
+
+\begin{lstlisting}
+
+sub foo {
+    bar(@_);
+}
+
+sub bar {
+   my ($a, %b) = @_;
+   baz($_[2]);
+   ...
+}
+
+sub baz {
+   close($_[0]) or die;
+}
+
+\end{lstlisting}
+
+All we can say with confidence (without seeing the remaining body of bar) is that the second argument to foo() is a hash which is likely only one key/value (as randomization will break this code), the value of which ought to be a filehandle, and which will throw in the event it isn't.
+
+Some tool which automates this reasoning, say a Sub::Characterizer would be of great utility. It could be used alongside Pod::Weaver annotations to largely auto-generate documentation as well.
+The core difficulty here is the sheer scope of such an undertaking.
+A lookup table of every call in core's arguments vis perlguts datatypes, along with every FFI interaction would have to be built, likely during BEGIN.
+From there, a set of boundary data for each of these datatypes would need to be maintained going forward.
+
+Some users of other programming languages believe themselves to have obviated the entire class of errors due to improper inputs via strong typing systems.
+This of course ignores the reality that poor choice of type (such as an "Any" type, or unsigned when you need signs, etc) immediately voids any such guarantees, and boundary testing is nonetheless necessary.
+
+b. But I've got 100% coverage, now what?
+
+It's usually fairly straightforward to get 100% statement coverage on greenfield code.
+However this rarely translates into software quality, as the test coverage on any of the deps in the stack remains opaque.
+How do we identify if the dependencies we use are well covered *where we use them*?
+If we can, the same techniques can be cross-referenced with the relative importance of the module vis the "CPAN River" to identify where the most high-impact test coverage improvements could be made in the perl ecosystem.
+
+It turns out that we can do this, supposing we actually run the testsuites of our dependencies when installing them.
+This is (normally) the case when using something like perlbrew, with an isolated environment.
+System perl is usually another story, as many modules will be shipped with another packaging manager.
+
+While we could always wrap CPAN, ideally a CPAN::Plugin either pre or post test would be ideal.
+Nevertheless, given the wide variety of means that CPAN Modules might be installed, any solution here would have to build a coverage index on-demand, and cache it for later use, likely as a plugin to Devel::Cover instead.
+cpancover *appears* to do this, but is currently broken and appears unmaintained.
+
+c. eval {} is not enough
+
+Since the primary programming paradigm is *not* "Klingon Programming" where it is better to die() than return() in failure, there is no guarantee that environmental problems are caught.
+This is why a great many Perl::Critic rules about using eval {...;1;} or do..., `my $pid=fork(); die unless defined $pid` and so on exist.
+Some means of actual fault injection should also be taken into account with regard to coverage.
+
+How many subs are actually safe when `ls` returns `read error`?
+Does your script recover from children whacked by oomkiller?
+What if fork() fails because the process table is filled with zombies?
+What about system() or qx{} or pipe open() when the target does not exist, but that -f returns true because you have a bad RAID card?
+
+This is actually a quite complicated subject.  As the founder of Anththesis (https://antithesis.com/blog/is_something_bugging_you/) puts it:
+
+the $100 bill is surrounded by an entire field of poisonous spike-pits, nay, a continent of them.
+
+Aside from building something which can inject the *astonishingly vast array* of such faults, there is still the problem of integrating this into perl's testing harness, and reporting coverage to do.
+
+d. Fitness for a particular purpose
+
+As mentioned earlier, the best means to evaluate whether changes to software harm it's fitness to purpose is some means of tracking user paths through the software as they perform tasks.
+There are two levels of this; one is front-end and the other on the back end.
+Ideally these are integrated somehow; given perl is not *usually* the code powering UIs, but generating them, our primary concern will be with the points that interfaces reach out and touch backend code.
+
+The best way to satisfy these needs is via a powerful logging framework using something like Log::Dispatch::DBI and robust session handling with shared identifiers between frontend and backend instrumentation.
+From there you need something to *identify and classify* common patterns of usage.
+This sounds suspiciously like what Neural Networks are *actually good at doing*, so let's do just that.
+
+e. Quality with non-functional Characteristics
+
+Devel::NYTProf is pretty much the best profiler in the industry; it's one of the primary reasons to recommend perl over alternatives.
+However, the ability to reason about the accessibility, localizability or portability of your program is crude at best.
+Generally the only option here is to impose linter rules which force the developer into the practices necessary to facilitate these things.
+The only systematic attempt to apply linting across the perl ecosystem is CPANTS, but this could stand to be extended with more categories of evaluation.
+
+For example, suppose I have the option of multiple modules on CPAN which I see do roughly the same thing in function X.
+It would greatly simplify my decision as to what to use if I already knew ahead of time:
+
+1. What's the bandwidth of this function; how does it perform?
+2. How complex is it?  I may have to read and debug this.
+3. Is it doing non-portable things?
+
+And so on.
+
+This is also another place where NNs could be of great utility; while they are pretty awful at writing code, they're pretty great at classifying it.
+As such we could likely build these lists of "Modules that do largely the same things on CPAN" in a mostly automated fashion.
+You could roll right up to metacpan and see a "More like this" link -- like youtube but actually useful!
+I for one would love seeing a table with these criteria and test coverage readily available -- no other packaging system does anything approaching this useful.
+
+Oftentimes your choice of components is the primary determinant of your output's quality; we could do a lot more to help people reason about this.
+
+\section{Conclusion}
+
+As you might expect from an active programming community of 35+ years, the things perl struggles to understand about its programs are essentially what every other programming language struggles with.
+This should come as no shock, as the ultimate cause of them are pretty sticky fundamental issues on the "even people don't qualify as Strong AI" spectrum.
+The good news is that our near-total incapacity to percieve all the ways in which our programs can fail also means automated solutions to ferreting these out are likely to be far more effective at doing so than users.
+As such the dream of "Bug Free" software is largely attainable in the same sense that a house in a bad neighborhood with cameras on it is going to be largely burgular free.
+Being of relatively higher quality than all your competitors is highly desirable, as quality is the *entire* driver of productivity and income.
+You don't have to be original, use non-dead programming languages or even be particularly talented to win with this approach. (japan.jpg)
+
+The software industry is one of the few places aside from SoEs (nasa.jpg) where this rate of failure is remotely tolerable.
+This is because our field is still relatively young, but that is changing quickly.
+The trend you need to get religion on is not AI, but quality.
+
 \end{document}

### 2024-05-21 sciperl/paper.tex
Yahoo, citations are working

@@
 \usepackage{graphicx}
+\usepackage{csquotes}
+\usepackage[T1]{fontenc}
 \usepackage[dvipsnames]{xcolor}
@@
 \lstset{language=Perl}
+
+\begin{filecontents}{paper.bib}
+    @misc{antithesis,
+        title   = "Is something Bugging you?",
+        author  = "Will Wilson",
+        url     = {https://antithesis.com/blog/is_something_bugging_you/},
+        year    = 2024
+    }
+\end{filecontents}
 
@@
 
+\newpage
 \section{INTRODUCTION}
@@
 
-1. rigorous testing of any individual *instruction* within a stack over it's expected domain of inputs, and a sampling of data outside of that is valuable.
+1. rigorous testing of any individual \textit{instruction} within a stack over it's expected domain of inputs, and a sampling of data outside of that is valuable.
 2.
@@
 Alternatively you can inspect the "intermediate form" which the interpreter itself builds via the B::Deparse, or similar techniques as Devel::Cover and Devel::NYTProf leverage.
-However just because you *can* reason about perl programs in the same fashion as any other programming language, does not guarantee avenues of inquiry are exhaustively explored.
+However just because you \textit{can} reason about perl programs in the same fashion as any other programming language, does not guarantee avenues of inquiry are exhaustively explored.
 
@@
 However this rarely translates into software quality, as the test coverage on any of the deps in the stack remains opaque.
-How do we identify if the dependencies we use are well covered *where we use them*?
-If we can, the same techniques can be cross-referenced with the relative importance of the module vis the "CPAN River" to identify where the most high-impact test coverage improvements could be made in the perl ecosystem.
+How do we identify if the dependencies we use are well covered \textit{where we use them}?
+If we can, the same techniques can be cross-referenced with the relative importance of the module vis.
+the "CPAN River" to identify where the most high-impact test coverage improvements could be made in the perl ecosystem.
 
@@
 Nevertheless, given the wide variety of means that CPAN Modules might be installed, any solution here would have to build a coverage index on-demand, and cache it for later use, likely as a plugin to Devel::Cover instead.
-cpancover *appears* to do this, but is currently broken and appears unmaintained.
+cpancover \textit{appears} to do this, but is currently broken and appears unmaintained.
 
@@
 
-Since the primary programming paradigm is *not* "Klingon Programming" where it is better to die() than return() in failure, there is no guarantee that environmental problems are caught.
+Since the primary programming paradigm is \textit{not} "Klingon Programming" (where it is better to \textit{die()} than \textit{return()} in failure), there is no guarantee that environmental problems are caught.
 This is why a great many Perl::Critic rules about using eval {...;1;} or do..., `my $pid=fork(); die unless defined $pid` and so on exist.
@@
 
-This is actually a quite complicated subject.  As the founder of Anththesis (https://antithesis.com/blog/is_something_bugging_you/) puts it:
-
-the $100 bill is surrounded by an entire field of poisonous spike-pits, nay, a continent of them.
-
-Aside from building something which can inject the *astonishingly vast array* of such faults, there is still the problem of integrating this into perl's testing harness, and reporting coverage to do.
+This is actually a quite complicated subject.  As the founder of Antithesis puts it \cite{antithesis}:
+\begin{displayquote}
+the \$100 bill is surrounded by an entire field of poisonous spike-pits, nay, a continent of them.
+\end{displayquote}
+Aside from building something which can inject the \textit{astonishingly vast array} of such faults, there is still the problem of integrating this into perl's testing harness, and reporting coverage to do.
 
@@
 
+\newpage
 \section{Conclusion}
@@
 
+\newpage
+\bibliographystyle{vancouver}
+\bibliography{paper}
+
 \end{document}

### 2024-05-21 sciperl/paper.tex
sectioned out

@@
-\documentclass[twoside]{article}
+\documentclass{article}
 \usepackage{hyperref}
@@
 From this two things are clear:
-
-1. rigorous testing of any individual \textit{instruction} within a stack over it's expected domain of inputs, and a sampling of data outside of that is valuable.
-2.
-The interaction of an instruction with its immediate follower and predecessor must also be well understood; these multi-instruction sections can indeed be treated as though they are themselves more complicated instructions.
-
+\begin{itemize}
+\item rigorous testing of any individual \textit{instruction} within a stack over it's expected domain of inputs, and a sampling of data outside of that is valuable.
+\item The interaction of an instruction with its immediate follower and predecessor must also be well understood; these multi-instruction sections can indeed be treated as though they are themselves more complicated instructions.
+\end{itemize}
 This is the actual theoretical underpinning behind the two testing concepts we are all familiar with (unit and integration testing), but most don't think too hard about.
@@
 
-a. Devel::Cover does not quite cover the subject
+\subsection{Devel::Cover does not quite cover the subject}
 
@@
 
-b. But I've got 100% coverage, now what?
+\subsection{But I've got 100\% coverage, now what?}
 
@@
 
-c. eval {} is not enough
+\subsection{eval \{\} is not enough}
 
@@
 
-d. Fitness for a particular purpose
+\subsection{Fitness for a particular purpose}
 
@@
 
-e. Quality with non-functional Characteristics
+\subsection{Quality with non-functional Characteristics}
 
@@
 It would greatly simplify my decision as to what to use if I already knew ahead of time:
-
-1. What's the bandwidth of this function; how does it perform?
-2. How complex is it?  I may have to read and debug this.
-3. Is it doing non-portable things?
-
+\begin{itemize}
+\item What's the bandwidth of this function; how does it perform?
+\item How complex is it?  I may have to read and debug this.
+\item Is it doing non-portable things?
+\end{itemize}
 And so on.

### 2024-05-21 sciperl/paper.tex
halfway done with citations

@@
 \begin{filecontents}{paper.bib}
+    @article{oring,
+        title = "The O-Ring Theory of Economic Development",
+        author = "Michael Kremer",
+        journal = "The Quarterly Journal of Economics, Vol. 108, No. 3",
+        url     = {https://www.jstor.org/stable/2118400},
+        year    = 1993,
+        note    = "I say 'most' here, because there is a not insignificant sector of the economy called the 'foolproof' sector, see Garret Jones' 2012 paper 'The O-Ring Sector and the Foolproof Sector'.
+However for the purposes of software, we can easily conclude much if not all of our industry is decidedly in the 'O-Ring' sector."
+    }
+    @misc{bezier,
+        title  = "Software testing techniques",
+        author = "Boris Bezier",
+        note    = "Chapter 1, Section 1.7 - ISBN:0442206720",
+        year    = 1990
+    }
+    @misc{bezier2,
+        title  = "ibid",
+        note   = "TODO: chapter/verse"
+    }
+    @misc{bezier3,
+        title  = "ibid",
+        note   = "Chapter 2, Section 3.4.1"
+    }
+    @misc{turing,
+        title  = "On Computable Numbers, with an Application to the Entscheidungsproblem",
+        url    = {http://www.turingarchive.org/browse.php/B/12},
+        author = "Alan Turing",
+        year   = 1937
+    }
+    @misc{frame,
+        title = "The Frame Problem",
+        url = {https://plato.stanford.edu/entries/frame-problem/},
+        note = "The current thinking as regards the resolution of the frame problem vis.
+AI is that intelligence, insofar as it exists at all, is 'embodied'.
+Which is to say it has physical presence in the world, and intelligence is drawn forth from application of the body's faculties thereupon.
+This explains a number of sociopolitical phenomenon such as the 'Flynn effect'."
+    }
+    @article{chapman,
+        title = "Planning for conjunctive goals",
+        journal = "Artificial Intelligence, Volume 32, Issue 3 pp. 333-377",
+        url = {https://www.sciencedirect.com/science/article/abs/pii/0004370287900920},
+        year = 1987,
+        author = "David Chapman",
+        note = "You can read it for free in MIT AI Technical Report 802 from 1985. Chapman proves NP-Completeness of the frame problem therein."
+    }
     @misc{antithesis,
@@
         url     = {https://antithesis.com/blog/is_something_bugging_you/},
-        year    = 2024
+        year    = 2024,
+        note    = "Having built similar systems myself, I second these sentiments."
     }
@@
 
-Productivity of any piece of capital or laborer is entirely a function of error rate; being able to quantify
-the conditions under which failures can be expected is thus of the utmost importance.
+The productivity of most capital or labor is \textit{almost entirely} a function of error rate \cite{oring}.
+Being able to quantify the conditions under which failures can be expected is thus of the utmost importance.
 
@@
 
-In software testing there's a concept known as the "pesticide paradox" which is shorthand for saying
+In software testing there's a concept known as the \textit{pesticide paradox} \cite{bezier} which is shorthand for saying
 "The bugs are always where you aren't looking".  Abscence of evidence is not evidence of abscence.
@@
 To identify where we are not looking, we need a proper mental model of how computers work.
-Thankfully, this is well established.  Programs are just stacks of instructions capable of adding and removing instructions to said stack.
+Thankfully, this is well established.
+Programs are just stacks of instructions capable of adding and removing instructions to said stack.
 Modern programs are usually multiple series of these stacks in parallel, usually referred to as pipelines.
@@
 
-Any given instruction is going to operate like a traditional mathematical function;
+Any given instruction is going to operate like any other mathematical function;
 this is to say that it will have a range of acceptable inputs and outputs.
@@
 \begin{itemize}
-\item rigorous testing of any individual \textit{instruction} within a stack over it's expected domain of inputs, and a sampling of data outside of that is valuable.
+\item rigorous testing of any individual \textit{instruction} within a stack over it's expected domain of inputs, and a sampling of data outside of that is valuable \cite{bezier2}.
 \item The interaction of an instruction with its immediate follower and predecessor must also be well understood; these multi-instruction sections can indeed be treated as though they are themselves more complicated instructions.
@@
 However the data you feed in is itself a stack of instructions with a necessarily unbounded input and output domain.
-This is best known as the "halting problem", as there is always theoretically some set of data you can feed a program which will prevent any program from ever terminating.
-What is less well understood is that the pesticide paradox is pernicious precisely because it is the halting problem.
-And the halting problem is actually Godel's Uncertainty principle - there will always exist at least one statement in any given axiomatic system which cannot ever be proven.
+To make things worse, well designed programs \textit{tend to} move code into data \cite{bezier3}, so now the problem not just undecidable \cite{turing} \textit{but also} recursive.
 
-This same problem rears its head when attempting to demonstrate fitness for a particular purpose!
-Any approach other than actually engaging in said purpose immediately falls afoul of the "frame problem".
-Ultimately nothing but an actual user can actually accept a product is fit for a particular purpose.
+Another vexatious problem rears its head when attempting to demonstrate fitness for a particular purpose!
+Any approach other than actually engaging in said purpose immediately falls afoul of the "frame problem" \cite{frame}.
+Ultimately nothing but an actual user can actually accept a product is fit for a particular purpose, because non-embodied intelligence can't do that \cite{chapman}.
 As such, any approach other than recording user journeys and being able to play them back is shooting in the dark.

### 2024-05-29 sciperl/paper.tex
first draft done

@@
 \usepackage[dvipsnames]{xcolor}
+\usepackage[english]{babel}
 \hypersetup{
@@
     }
+    @misc{prodprop,
+        note = "While this was evident to even the early Taylorists, the idea did not become widely popular before Japanese adoption of Deming and Juran's principles."
+    }
     @misc{bezier2,
-        title  = "ibid",
+        title  = "Bezier's software testing techniques",
         note   = "TODO: chapter/verse"
@@
         note    = "Having built similar systems myself, I second these sentiments."
+    }
+    @misc{mises1,
+        title = "Human Action",
+        author = "Ludwig Von Mises",
+        url    = {https://mises.org/library/book/human-action},
+        year   = 1949,
+        note   = "Ch. I, Sec IV"
+    }
+    @misc{mises2,
+        title = "ibid",
+        note  = "Ch. XXVI"
+    }
+    @misc{atwood,
+        title  = "The God Login",
+        author = "Jeff Atwood",
+        url    = {https://blog.codinghorror.com/the-god-login/},
+        year   = 2015,
+        note   = "While I've heard many testers describe such comprehensive user surveillance systems they've built over the years, I've never heard them classified.
+Perhaps the appropriate name might be 'revelation' tests."
+    }
+    @misc{schlep,
+        title  = "Schlep Blindness",
+        author = "Paul Graham",
+        url    = {https://paulgraham.com/schlep.html},
+        year   = 2012
+    }
+    @misc{gsb,
+        url  = {https://github.com/teodesian/Pod-Coverage-More/blob/master/lib/Pod/Coverage/More.pm}
+    }
+    @misc{gsb2,
+        note = "I once built a test script for just this, and then used the linux kernel testing api to put the zombies in uninterruptible sleep().  This is a fun way to DOS hypervisors."
+    }
+    @misc{mock,
+        url = {https://metacpan.org/pod/Test::MockFile#How-this-mocking-is-done:}
+    }
+    @article{copilot,
+        title="Measuring Github Copilot's Impact on Productivity",
+        journal="Communications of the ACM",
+        year=2024,
+        author="Albert Ziegler",
+        author="Eirini Kalliamvakou",
+        author="X. Alice Li",
+        author="Andrew Rice",
+        author="Devon Rifkin",
+        author="Shawn Simister",
+        author="Ganesh Sittampalam",
+        author="Edward Aftandilian",
+        url={https://cacm.acm.org/research/measuring-github-copilots-impact-on-productivity/},
+        note="'For context, in our study GitHub Copilot has an acceptance rate of 27\%'.
+While I've worked with some bad programmers (matthew 7:3), I've never worked long with ones who required 4 rounds of review to even be close to correct."
+    }
+    @book{standish,
+        author="Jim Jordan, Standish Group",
+        year=2021,
+        title="CHAOS report: Beyond Infinity",
+        note="Like Brooks' 'Mythical Man Month' predicted, it has borne out that software failure rates remain stubbornly high."
     }
@@
 
-To identify where we are not looking, we need a proper mental model of how computers work.
+\subsection{Fundamental Testing Principles and Problems}
+
+To identify where we are not looking, we need a proper mental model of how computers work if we are to test them.
 Thankfully, this is well established.
@@
 Much like a child's game of "telephone", the earlier in the stack this happens, typically the worse the outcome.
+Professional software testers are quick to tell you that this is \textit{also} true of the software \textit{production process} itself \cite{prodprop}.
 
@@
 
-Another vexatious problem rears its head when attempting to demonstrate fitness for a particular purpose!
+\subsection{Fitness: a further complication}
+
+Another vexatious problem rears its head when attempting to demonstrate fitness for a particular purpose.
 Any approach other than actually engaging in said purpose immediately falls afoul of the "frame problem" \cite{frame}.
@@
 
-Economists call this concept "Demonstrated Preference", and the impossibility of centralized economic planning is largely the frame problem restated.
-Many firms fall victim to this conceit with their "user stories" and acceptance tests based around these.
-Ultimately the only means to catch fitness issues early in new features is to eat your own dog food, being particularly attentive to "Schlepping" and other places the work-flow can be de-burred.
+Economists call this concept "Demonstrated Preference" and "Subjective Value"\cite{mises1},
+and it is my hypothesis that the impossibility of centralized economic planning \cite{mises2} is largely the frame problem restated.
+For our purposes, it is sufficient to say that while computers are quite good at dealing with cardinal concepts, they are astonishingly poor at dealing with ordinal ones.
+Anyone familiar with sorting algorithms, counting problems and the "perfect covering" problem will have figured this out the hard way.
+In this case, not only are we dealing with a problem which would be at least NP-Hard had we perfect knowledge, we also must grapple with incomplete information.
+
+We are left with nothing more than the "God Algorithm" (already know the answer). \cite{atwood}
+Frequently this is how caching systems are described, but to some extent this is also true of user surveillance.
+It should therefore come as no surprise that tools enabling comprehensive user surveillance have proven quite popular.
+
+Nevertheless, many firms fall victim to the conceit of planning with their "user stories" and acceptance tests based around these.
+Ultimately the only \textit{effective} means to catch fitness issues early in new features is to "eat your own dog food",
+being particularly attentive to "Schleps" \cite{schlep} and other places the work-flow can be de-burred or otherwise streamlined.
 
@@
 
-All programs can after all be expressed as lambda calculus; programs are nearly always transformed into some variety of "intermediate form" prior to being handed off to lower level executors.
+Knowing all this, what then shall we do?
+First, let's go over the extant tools on CPAN.
+
+All programs can be expressed as lambda calculus; programs are nearly always transformed into some variety of "intermediate form" prior to being handed off to lower level executors.
 One such form is the "abstract syntax tree", which is an intermediate form used to reason about the correctness of programs rather than to pass to hardware for execution.
@@
 Devel::Cover is quite good at identifying statements and conditions which are not covered within a program.
-It cannot however identify places where a condition ought to be applied, but is not.
+It cannot however identify places where a condition ought to be applied, but is not or vice versa.
 
@@
 It is generally via these means that we discover these missed conditionals, as we are frequently surprised by the realities imposed by input data.
-Given perl only has optional input signatures, and no output signatures nor ways to signify that a sub throws, this has to be inferred; a characterization of the stack must be built.
+Given perl only has optional input signatures, and no output signatures nor ways to signify that a sub throws, this has to be inferred;
+a characterization of the stack must be built.
 
@@
 
-All we can say with confidence (without seeing the remaining body of bar) is that the second argument to foo() is a hash which is likely only one key/value (as randomization will break this code), the value of which ought to be a filehandle, and which will throw in the event it isn't.
+All we can say with confidence (without seeing the remaining body of bar) is that the second argument to foo() is a hash which is likely only one key/value (as randomization will break this code),
+the value of which ought to be a filehandle, and which will throw in the event it isn't.
 
-Some tool which automates this reasoning, say a Sub::Characterizer would be of great utility. It could be used alongside Pod::Weaver annotations to largely auto-generate documentation as well.
+Some tool which automates this reasoning, say a Sub::Characterizer would be of great utility.
+It could be used alongside Pod::Weaver annotations to largely auto-generate documentation as well.
+Pod::Coverage could be similarly enhanced.
+
 The core difficulty here is the sheer scope of such an undertaking.
@@
 From there, a set of boundary data for each of these datatypes would need to be maintained going forward.
+Even then you'll want some means of annotating when a sub's arguments acceptable input domain is a subset of its governing datatype.
+I found most of this out the "hard way" \cite{gsb} long ago attempting to solve this particular problem (and failing).
+
+\subsection{Surely strong typing solves all this?}
 
 Some users of other programming languages believe themselves to have obviated the entire class of errors due to improper inputs via strong typing systems.
-This of course ignores the reality that poor choice of type (such as an "Any" type, or unsigned when you need signs, etc) immediately voids any such guarantees, and boundary testing is nonetheless necessary.
+This of course ignores the reality that poor choice of type (such as an "Any" type, or unsigned when you need signs, etc) immediately voids any such guarantees.
+Boundary testing also remains necessary.
+However the argument is not entirely without merit, and there are some things such as the `types` pragma and various other runtime type-checkers such as Type::Tiny to attest to this.
 
-\subsection{But I've got 100\% coverage, now what?}
+Typescript (a code generator) is probably the best example of a dynamic language figuring out how to give you such compile-time guarantees.
+Supposing everything else it interacts with is also generated by typescript, a good degree of confidence can be had.
+It's also likely that much of the code therein, particularly the types themselves, can be repurposed towards doing the same with perl.
+While I wished to build a proof of concept for this, alas time did not permit.
+Building a generic framework to generate dynamic language code from typed variants might be an incredibly good fit for Raku Grammars.
 
-It's usually fairly straightforward to get 100% statement coverage on greenfield code.
-However this rarely translates into software quality, as the test coverage on any of the deps in the stack remains opaque.
+\subsection{I've got 100\% coverage though, its perfect right?}
+
+It's usually fairly straightforward to get 100\% statement coverage on greenfield code.
+However this rarely translates into software quality, as the test coverage on any of the dependencies in the stack is opaque in practically all programming languages.
 How do we identify if the dependencies we use are well covered \textit{where we use them}?
@@
 
+Yet another place where I would love to have a proof of concept.
+
 \subsection{eval \{\} is not enough}
 
-Since the primary programming paradigm is \textit{not} "Klingon Programming" (where it is better to \textit{die()} than \textit{return()} in failure), there is no guarantee that environmental problems are caught.
-This is why a great many Perl::Critic rules about using eval {...;1;} or do..., `my $pid=fork(); die unless defined $pid` and so on exist.
-Some means of actual fault injection should also be taken into account with regard to coverage.
+Since not all programmers adhere to "Klingon Programming" (where it is better to \textit{die()} than \textit{return()} in failure),
+there is no guarantee that environmental problems are caught.
+This is why a great many Perl::Critic rules exist to catch precisely such problems caused by external factors.
+For example:
+\begin{lstlisting}
+eval {...;1;} or do...
+
+my $pid=fork(); die unless defined $pid;
+\end{lstlisting}
+Clearly, some means of actual fault injection should also be taken into account with regard to coverage.
 
@@
 Does your script recover from children whacked by oomkiller?
-What if fork() fails because the process table is filled with zombies?
-What about system() or qx{} or pipe open() when the target does not exist, but that -f returns true because you have a bad RAID card?
+What if fork() fails because the process table is filled with zombies? \cite{gsb2}
+What about when system() or qx{} or pipe open() where the target does not exist, but -f returns true because you have a bad RAID card?
+I (and I suspect many of you) have had the misfortune of experiencing these cases personally.
 
@@
 \end{displayquote}
-Aside from building something which can inject the \textit{astonishingly vast array} of such faults, there is still the problem of integrating this into perl's testing harness, and reporting coverage to do.
+Aside from building something which can inject the \textit{astonishingly vast array} of such faults,
+there is still the problem of integrating this into perl's testing harness, and reporting coverage of such in some fashion.
+
+Given the scope, I have made no attempts at a proof of concept.
+However, if I were to do so, I think the approach used by Test::MockFile \cite{mf} is likely one which can be leveraged to satisfy these needs.
 
@@
 As mentioned earlier, the best means to evaluate whether changes to software harm it's fitness to purpose is some means of tracking user paths through the software as they perform tasks.
+If sudden unintended disturbances in user paths are detected after pushing changes, it is a strong indication of error.
 There are two levels of this; one is front-end and the other on the back end.
-Ideally these are integrated somehow; given perl is not *usually* the code powering UIs, but generating them, our primary concern will be with the points that interfaces reach out and touch backend code.
+Ideally these are integrated somehow; given perl is not \textit{usually} the code powering UIs, but generating them, our primary concern will be with the points that interfaces reach out and touch backend code.
 
 The best way to satisfy these needs is via a powerful logging framework using something like Log::Dispatch::DBI and robust session handling with shared identifiers between frontend and backend instrumentation.
-From there you need something to *identify and classify* common patterns of usage.
-This sounds suspiciously like what Neural Networks are *actually good at doing*, so let's do just that.
+From there you need something to \textit{identify and classify} common patterns of usage.
+This sounds suspiciously like what Neural Networks are \textit{actually good at doing}.
+
+AI::Categorizer running on a pack()ed representation of an application's state parameters throughout a user session ought to do the trick.
+Yet another place I will have to rush to get a POC if I can at all.
 
@@
 
-This is also another place where NNs could be of great utility; while they are pretty awful at writing code, they're pretty great at classifying it.
+This is also another place where NNs could be of great utility; while they are necessarily awful at writing \textit{correct} code \cite{copilot}, they're pretty great at classifying it.
 As such we could likely build these lists of "Modules that do largely the same things on CPAN" in a mostly automated fashion.
@@
 As you might expect from an active programming community of 35+ years, the things perl struggles to understand about its programs are essentially what every other programming language struggles with.
-This should come as no shock, as the ultimate cause of them are pretty sticky fundamental issues on the "even people don't qualify as Strong AI" spectrum.
+This should come as no shock, as the ultimate cause of them are pretty sticky fundamental issues of the "even people don't qualify as Strong AI" stripe.
 The good news is that our near-total incapacity to percieve all the ways in which our programs can fail also means automated solutions to ferreting these out are likely to be far more effective at doing so than users.
 As such the dream of "Bug Free" software is largely attainable in the same sense that a house in a bad neighborhood with cameras on it is going to be largely burgular free.
-Being of relatively higher quality than all your competitors is highly desirable, as quality is the *entire* driver of productivity and income.
-You don't have to be original, use non-dead programming languages or even be particularly talented to win with this approach. (japan.jpg)
+Being of relatively higher quality than all your competitors is highly desirable, as quality is the \textit{entire} driver of productivity and income.
+You don't have to be original, use "alive" programming languages or even be particularly talented to win with this approach.
 
-The software industry is one of the few places aside from SoEs (nasa.jpg) where this rate of failure is remotely tolerable.
-This is because our field is still relatively young, but that is changing quickly.
-The trend you need to get religion on is not AI, but quality.
+The software industry is one of the few places aside from SoEs where a 2/3 rate of failure \cite{standish} is remotely tolerable.
+This is because our field is still relatively young (and thus high margin), but that is changing quickly.
+The trend you need to get behind is not AI, but quality.
+Whoever in our industry fully internalizes the lessons of Fred Brooks, Deming and Juran first will win.
+Tooling such as suggested here would greatly further those ends, and give whatever language that builds them great advantage over their alternatives.
 

### 2024-06-04 sciperl/paper.tex
yet more

@@
 
-In the era of testsuite generation via large language models, this is of even greater concern.
-
 I will lay out what is currently available, how that fits into a framework for reasoning about programs
@@
 Anything outside of these ranges results in either invalid or undefined behavior when composed in a stack of instructions.
-Much like a child's game of "telephone", the earlier in the stack this happens, typically the worse the outcome.
+Much like a child's game of telephone\footnote{Also known as chinese whispers, and other names}, the earlier in the stack this happens, typically the worse the outcome.
 Professional software testers are quick to tell you that this is \textit{also} true of the software \textit{production process} itself \cite{prodprop}.
@@
 \end{itemize}
-This is the actual theoretical underpinning behind the two testing concepts we are all familiar with (unit and integration testing), but most don't think too hard about.
-It's this lack of concrete understanding that leads to frequent confusion and ineffective testsuites the world over.
+These are the concepts behind two of the three\footnote{the third being acceptance or fitness testing which is discussed later} approaches to testing programs (unit\footnote{also referred to component testing} and integration\footnote{frequently confused with acceptance and black or grey-box testing} testing).
 
-It is tempting from here to consider programs as discrete stacks, which can be exhaustively proven to behave well, in a sort of "proof by induction".
+It is tempting from here to consider programs as discrete stacks, which can be exhaustively proven to behave well, in a \textit{proof via induction}.
 However the data you feed in is itself a stack of instructions with a necessarily unbounded input and output domain.
@@
 
-Another vexatious problem rears its head when attempting to demonstrate fitness for a particular purpose.
+Another vexatious problem is encountered when demonstrating fitness for a particular purpose.
 Any approach other than actually engaging in said purpose immediately falls afoul of the "frame problem" \cite{frame}.
@@
 
-Economists call this concept "Demonstrated Preference" and "Subjective Value"\cite{mises1},
-and it is my hypothesis that the impossibility of centralized economic planning \cite{mises2} is largely the frame problem restated.
-For our purposes, it is sufficient to say that while computers are quite good at dealing with cardinal concepts, they are astonishingly poor at dealing with ordinal ones.
-Anyone familiar with sorting algorithms, counting problems and the "perfect covering" problem will have figured this out the hard way.
+Economists call this concept "Demonstrated Preference" and "Subjective Value"\cite{mises1}\footnote{it is my hypothesis that the impossibility of centralized economic planning \cite{mises2} is largely the frame problem restated.}.
+For our purposes, it is sufficient to say that while computers are quite good at dealing with cardinal\footnote{a member of a cardinal set is \textit{equal, greater or lesser} than another} concepts,
+they are astonishingly poor at dealing with ordinal\footnote{a member of an ordinal set is \textit{ordered before or after} another; X is preferred to Y.
+Duplicates may effect said ordering based on a \textit{marginal utility function}.} ones.
+Anyone familiar with sorting algorithms, counting problems and Ramsey's\footnote{Also known as the clique problem.
+See here for my latest brush with this: https://troglodyne.net/posts/6ff13eb0-13de-11ec-84d9-8f4e0c69f986} problem will have figured this out the hard way.
 In this case, not only are we dealing with a problem which would be at least NP-Hard had we perfect knowledge, we also must grapple with incomplete information.
@@
 Frequently this is how caching systems are described, but to some extent this is also true of user surveillance.
-It should therefore come as no surprise that tools enabling comprehensive user surveillance have proven quite popular.
+This is one of many reasons that tools enabling comprehensive user surveillance have proven quite popular\footnote{Mostly for the purpose of advertisement}.
 
-Nevertheless, many firms fall victim to the conceit of planning with their "user stories" and acceptance tests based around these.
-Ultimately the only \textit{effective} means to catch fitness issues early in new features is to "eat your own dog food",
-being particularly attentive to "Schleps" \cite{schlep} and other places the work-flow can be de-burred or otherwise streamlined.
+Nevertheless, many firms fall victim to the conceit of planning with \textit{user stories}\footnote{An idealized conception of a user and how their motivations might propel their journey through a usage session} and acceptance tests based around these.
+Ultimately the only \textit{effective} means to catch fitness issues early in new features is to \textit{eat your own dog food}\footnote{Origins of the phrase are largely lost to time, but the practice was generally accepted in American firms before the 1990s software boom.},
+being particularly attentive to "Schleps" \cite{schlep} and other places the work-flow can be de-burred or otherwise streamlined\footnote{The most popular systematic approaches to this being Kaizen and Kanban}.
 
@@
 Knowing all this, what then shall we do?
-First, let's go over the extant tools on CPAN.
+First, let's go over the extant tools on CPAN\footnote{Comprehensive Perl Archive network.  https://cpan.org}.
 
@@
 
-Building such a tree is how PPI, and the tools which rely on it such as Perl::Critic work.
-Alternatively you can inspect the "intermediate form" which the interpreter itself builds via the B::Deparse, or similar techniques as Devel::Cover and Devel::NYTProf leverage.
-However just because you \textit{can} reason about perl programs in the same fashion as any other programming language, does not guarantee avenues of inquiry are exhaustively explored.
+Building such a tree is how PPI\footnote{An important distinction must be made here: the intermediate form is of Perl \textit{documents} rather than Perl \textit{programs}; which is to say an idealized conception of Perl which does not necessarily line up with interpreted behavior, but is close enough to be useful.},
+and the tools which rely on it such as \texttt{Perl::Critic}\footnote{https://metacpan.org/pod/Perl::Critic} work.
+Alternatively you can inspect the "intermediate form" which the interpreter itself builds via the \texttt{B::Deparse}\footnote{https://metacpan.org/pod/B::Deparse},
+or similar techniques such as \texttt{Devel::Cover}\footnote{https://metacpan.org/pod/Devel::Cover} and \texttt{Devel::NYTProf}\footnote{https://metacpan.org/pod/Devel::NYTProf} leverage.
+However the ability to reason about Perl programs in the same fashion as any other programming language does not guarantee our avenues of inquiry are exhaustively explored.
 
@@
 
-Devel::Cover is quite good at identifying statements and conditions which are not covered within a program.
+\texttt{Devel::Cover} is quite good at identifying statements and conditions which are not covered within a program.
 It cannot however identify places where a condition ought to be applied, but is not or vice versa.
@@
 It also does not have a means of identifying whether the boundaries of any given block's input/output domain have been exercised.
-It is generally via these means that we discover these missed conditionals, as we are frequently surprised by the realities imposed by input data.
-Given perl only has optional input signatures, and no output signatures nor ways to signify that a sub throws, this has to be inferred;
-a characterization of the stack must be built.
+It is generally via testing these boundaries that needs for additional conditional statements\footnote{Sometimes referred to guard clauses, short-circuits, etc} are discovered.
+Unfortunately the means available to automatically discover these boundaries is lacking.
+
+Perl only has optional input signatures\footnote{https://perldoc.pl/perlsub\#Signatures}, and no output signatures.
+Nor are there ways to signify that a subroutine throws exceptions.
+All this must be inferred via some means.
 
@@
 
-All we can say with confidence (without seeing the remaining body of bar) is that the second argument to foo() is a hash which is likely only one key/value (as randomization will break this code),
-the value of which ought to be a filehandle, and which will throw in the event it isn't.
+There is little we can say with confidence here.
+What we can say (without seeing the remaining body of bar) is that:
+\begin{itemize}
+\item The second argument to foo() is a likely a hash due to assignment (but might not be, as it is then later called-by-reference!)
+\item This hash is (hopefully) one key/value (as randomization will break this code due to aforementioned call-by-reference of the third argument which is presumably the hash value)
+\item This value ought to be an \textit{open} filehandle, given it is then passed to close()
+\item In the event it is not, the subroutine throws an exception.
+\end{itemize}
+As you can see, there is \textit{considerable} ambiguity in even quite trivial examples.
 
-Some tool which automates this reasoning, say a Sub::Characterizer would be of great utility.
-It could be used alongside Pod::Weaver annotations to largely auto-generate documentation as well.
-Pod::Coverage could be similarly enhanced.
+Some tool which automates this reasoning, likely a CPAN module (Which might appropriately be named \texttt{Sub::Characterizer}), would be of great utility.
+It could be used alongside \texttt{Pod::Weaver}\footnote{https://metacpan.org/pod/Pod::Weaver} annotations to largely auto-generate subroutine documentation as well.
+\texttt{Pod::Coverage}\footnote{https://metacpan.org/pod/Pod::Coverage} could be similarly enhanced.
 
-The core difficulty here is the sheer scope of such an undertaking.
-A lookup table of every call in core's arguments vis perlguts datatypes, along with every FFI interaction would have to be built, likely during BEGIN.
-From there, a set of boundary data for each of these datatypes would need to be maintained going forward.
-Even then you'll want some means of annotating when a sub's arguments acceptable input domain is a subset of its governing datatype.
+The difficulty here is the scope of such an undertaking.
+A lookup table of every subroutine in Perl's CORE\footnote{\texttt{perldoc perlfunc}} namespace's arguments mapping to perl's internal datatypes \footnote{see \texttt{perldoc perlguts}} would need to be built.
+The same would have to be done for every FFI\footnote{Foreign Function Interface} or XS\footnote{\texttt{perldoc perlxs}} module used.
+Given all other Perl subroutines you write are going to be a composition thereof, this knowledge would allow for correct characterizations of subroutines to be made.
+
+There are \textit{significant} costs involved in this, particularly for FFI/XS, which is much of CPAN.
+This is because the FFI interface itself (be it XS, libffi\footnote{the backend for \texttt{FFI::Platypus}}, or others) would have to produce this information as a build artifact
+which could then be used to instruct the characterizer appropriately.
+
+In addition, a set of boundary data for each of these datatypes would need to be maintained going forward, along with whether fatal exceptions may occur.
+Even then it would be of great utility to have some means of annotating when a subroutine's arguments acceptable input domain is a subset of its governing datatype.
 I found most of this out the "hard way" \cite{gsb} long ago attempting to solve this particular problem (and failing).
@@
 Boundary testing also remains necessary.
-However the argument is not entirely without merit, and there are some things such as the `types` pragma and various other runtime type-checkers such as Type::Tiny to attest to this.
+However the argument is not entirely without merit, and there are various runtime constraint systems such as \texttt{Type::Tiny} attesting to this\footnote{Along with other efforts, such as those promoted by the Perl Types committee: https://perlcommunity.org/types/}.
 
-Typescript (a code generator) is probably the best example of a dynamic language figuring out how to give you such compile-time guarantees.
-Supposing everything else it interacts with is also generated by typescript, a good degree of confidence can be had.
+Typescript\footnote{a ECMAScript (aka Javascript) code generator} is probably the best example of a dynamic language figuring out how to give you such compile-time guarantees.
+Supposing everything else it interacts with is also generated by Typescript\footnote{Rarely ever the case in Javascript projects}, a good degree of confidence can be had.
 It's also likely that much of the code therein, particularly the types themselves, can be repurposed towards doing the same with perl.
-While I wished to build a proof of concept for this, alas time did not permit.
-Building a generic framework to generate dynamic language code from typed variants might be an incredibly good fit for Raku Grammars.
 

### 2024-06-04 sciperl/paper.tex
largely good

@@
         note = "You can read it for free in MIT AI Technical Report 802 from 1985. Chapman proves NP-Completeness of the frame problem therein."
-    }
-    @misc{antithesis,
-        title   = "Is something Bugging you?",
-        author  = "Will Wilson",
-        url     = {https://antithesis.com/blog/is_something_bugging_you/},
-        year    = 2024,
-        note    = "Having built similar systems myself, I second these sentiments."
     }
@@
 
-\subsection{I've got 100\% coverage though, its perfect right?}
+\subsection{100\% coverage, or: how many angels can dance upon the head of a pin?}
 
-It's usually fairly straightforward to get 100\% statement coverage on greenfield code.
-However this rarely translates into software quality, as the test coverage on any of the dependencies in the stack is opaque in practically all programming languages.
+It's usually fairly straightforward to get 100\% statement coverage on greenfield\footnote{ex nihilo} code.
+However this rarely translates into software quality, as the test coverage for any of the dependencies in the stack is opaque in practically all programming languages.
 How do we identify if the dependencies we use are well covered \textit{where we use them}?
-If we can, the same techniques can be cross-referenced with the relative importance of the module vis.
-the "CPAN River" to identify where the most high-impact test coverage improvements could be made in the perl ecosystem.
 
-It turns out that we can do this, supposing we actually run the testsuites of our dependencies when installing them.
+First, we must run the testsuites of our dependencies when installing them.
 This is (normally) the case when using something like perlbrew, with an isolated environment.
@@
 While we could always wrap CPAN, ideally a CPAN::Plugin either pre or post test would be ideal.
-Nevertheless, given the wide variety of means that CPAN Modules might be installed, any solution here would have to build a coverage index on-demand, and cache it for later use, likely as a plugin to Devel::Cover instead.
-cpancover \textit{appears} to do this, but is currently broken and appears unmaintained.
+Given the wide variety of means that CPAN Modules might be installed, any solution here would have to build a coverage index on-demand (likely caching it for later use).
+cpancover \textit{appears} it \textit{intends to} do this, but is currently broken and appears unmaintained.
+
+The same techniques can be cross-referenced with the relative importance of the module.
+This information is known as the \textit{CPAN River}\footnote{See http://neilb.org/2015/04/20/river-of-cpan.html}.
+Doing this will identify where the most high-impact test coverage improvements can be made in the perl ecosystem.
 
@@
 
-Since not all programmers adhere to "Klingon Programming" (where it is better to \textit{die()} than \textit{return()} in failure),
-there is no guarantee that environmental problems are caught.
-This is why a great many Perl::Critic rules exist to catch precisely such problems caused by external factors.
-For example:
+Since not all subroutines are \textit{Klingon Programming}\footnote{where it is better to \textit{die()} than \textit{return()} in failure.
+Also known as "Control-Flow by Exception".
+Such subroutines usually behave poorly in threaded contexts.},
+such failures tend to go unnoticed when inured in this mental model.
+This is why a great many Perl::Critic policies exist to catch precisely such problems caused by external factors largely outside of a subroutine's control.
+For example, I have observed policies enforcing code along these lines:
 \begin{lstlisting}
-eval {...;1;} or do...
+# Failing to have a 'do' clause after block eval without
+# assignment generally means something is failing silently
+eval {...;1;} or do {...}
 
-my $pid=fork(); die unless defined $pid;
+# Failing to check that you got a defined pid from fork()
+# means the program will have undefined behavior in the
+# event the process table is full.
+my $pid=fork() // die;
 \end{lstlisting}
-Clearly, some means of actual fault injection should also be taken into account with regard to coverage.
+Some means of actual fault injection should also be taken into account in addition to standard coverage metrics as a measure of the code's overall robustness.
 
-How many subs are actually safe when `ls` returns `read error`?
-Does your script recover from children whacked by oomkiller?
-What if fork() fails because the process table is filled with zombies? \cite{gsb2}
-What about when system() or qx{} or pipe open() where the target does not exist, but -f returns true because you have a bad RAID card?
-I (and I suspect many of you) have had the misfortune of experiencing these cases personally.
+The sources of these faults outside of a subroutine's control are generally syscalls or FFI.
+As such we will need some means of discerning whether or not the usage in a test behaves well under these scenarios:
+\begin{itemize}
+\item call emits (or is subjected to) a signal (e.g. SIGUSR1, SIGHUP, SIGKILL) either directly or in a child
+\item call emits an exception
+\item call returns a failure code
+\item call appears successful, but nonetheless failed
+\end{itemize}
+Ideally, this would be done in a semi-automated fashion where one could write a test exercising the expected operational modes, and then re-execute subjected to permutations of these failures.
+The approach used by \texttt{Test::MockFile}\cite{mock} is likely one which can be leveraged to inject syscall faults, and everything else can be handled with a universal mock (which would wrap whatever existing mock or sub is called.).
 
-This is actually a quite complicated subject.  As the founder of Antithesis puts it \cite{antithesis}:
-\begin{displayquote}
-the \$100 bill is surrounded by an entire field of poisonous spike-pits, nay, a continent of them.
-\end{displayquote}
-Aside from building something which can inject the \textit{astonishingly vast array} of such faults,
-there is still the problem of integrating this into perl's testing harness, and reporting coverage of such in some fashion.
+Automatically identifying all the subroutine calls within a given subroutine to inject these faults is complicated and relatively costly.
+A look-ahead using a similar technique as used in \texttt{Devel::Trace}\footnote{https://metacpan.org/pod/Devel::Trace} (or internal interpreter knowledge) would need to be used to know whether we are in the right place.
+Furthermore, we must pick the subroutines out of the next line if we are in the right place and wrap/mock them appropriately.
 
-Given the scope, I have made no attempts at a proof of concept.
-However, if I were to do so, I think the approach used by Test::MockFile \cite{mf} is likely one which can be leveraged to satisfy these needs.
+Thankfully there is a solution which uses XS to accomplish this: \texttt{Devel::Probe}\footnote{https://metacpan.org/pod/Devel::Probe}.
+
+TODO: show off POC doing this with tests...
 
@@
 If sudden unintended disturbances in user paths are detected after pushing changes, it is a strong indication of error.
-There are two levels of this; one is front-end and the other on the back end.
-Ideally these are integrated somehow; given perl is not \textit{usually} the code powering UIs, but generating them, our primary concern will be with the points that interfaces reach out and touch backend code.
 
-The best way to satisfy these needs is via a powerful logging framework using something like Log::Dispatch::DBI and robust session handling with shared identifiers between frontend and backend instrumentation.
+This is generally done via coupling a powerful logging framework such as \texttt{Log::Dispatch::DBI}\footnote{https://metacpan.org/pod/Log::Dispatch::DBI} and robust session handling using unique identifiers.
 From there you need something to \textit{identify and classify} common patterns of usage.
@@
 
-AI::Categorizer running on a pack()ed representation of an application's state parameters throughout a user session ought to do the trick.
-Yet another place I will have to rush to get a POC if I can at all.
+\texttt{AI::Categorizer} running on a \texttt{pack()}ed representation of an application's state parameters throughout a user session ought to do the trick.
 
@@
 
-Devel::NYTProf is pretty much the best profiler in the industry; it's one of the primary reasons to recommend perl over alternatives.
+Tools for reasoning about the performance of perl programs such as \texttt{Devel::NYTProf}\footnote{https://metacpan.org/pod/Devel::NYTProf} cover the subject of performance well.
 However, the ability to reason about the accessibility, localizability or portability of your program is crude at best.
 Generally the only option here is to impose linter rules which force the developer into the practices necessary to facilitate these things.
-The only systematic attempt to apply linting across the perl ecosystem is CPANTS, but this could stand to be extended with more categories of evaluation.
+The only systematic attempt to apply such linting across the perl ecosystem is CPANTS\footnote{https://cpants.cpanauthors.org/},
+but this could stand to be extended with more categories of evaluation.
 
@@
 \begin{itemize}
-\item What's the bandwidth of this function; how does it perform?
+\item how does it perform?
 \item How complex is it?  I may have to read and debug this.
@@
 
-This is also another place where NNs could be of great utility; while they are necessarily awful at writing \textit{correct} code \cite{copilot}, they're pretty great at classifying it.
+This is also another place where AI techniques could be of great utility.
+While they are necessarily awful at writing \textit{correct} code \cite{copilot}, they're pretty great at classifying code.
 As such we could likely build these lists of "Modules that do largely the same things on CPAN" in a mostly automated fashion.
-You could roll right up to metacpan and see a "More like this" link -- like youtube but actually useful!
-I for one would love seeing a table with these criteria and test coverage readily available -- no other packaging system does anything approaching this useful.
+Currently, the best we have is the "SEE ALSO" section of a module's POD, which is far from comprehensive.
 
-Oftentimes your choice of components is the primary determinant of your output's quality; we could do a lot more to help people reason about this.
+Coupling CPANTS to a classifier would make it far more useful, stimulating the development of these additional checks which do not currently exist.
+Being able to sort a list of candidate modules for a particular purpose by \texttt{kwalitee}\footnote{CPANTS refers to non-functional parameters as kwalitee}
+is one way to accomplish that.
 
@@
 
-As you might expect from an active programming community of 35+ years, the things perl struggles to understand about its programs are essentially what every other programming language struggles with.
-This should come as no shock, as the ultimate cause of them are pretty sticky fundamental issues of the "even people don't qualify as Strong AI" stripe.
-The good news is that our near-total incapacity to percieve all the ways in which our programs can fail also means automated solutions to ferreting these out are likely to be far more effective at doing so than users.
-As such the dream of "Bug Free" software is largely attainable in the same sense that a house in a bad neighborhood with cameras on it is going to be largely burgular free.
-Being of relatively higher quality than all your competitors is highly desirable, as quality is the \textit{entire} driver of productivity and income.
-You don't have to be original, use "alive" programming languages or even be particularly talented to win with this approach.
+As you might expect from an active programming community of 35+ years, the things Perl programmers struggle to understand about their programs are what every programming language community struggles with.
+The good news is that our incapacity to percieve all the ways in which our programs can fail also means automated solutions to ferreting these out are likely to be far more effective at doing so than users.
+Much of this tooling has yet to have been built, and recent advances have put a number of them tantalizingly within our grasp.
 
-The software industry is one of the few places aside from SoEs where a 2/3 rate of failure \cite{standish} is remotely tolerable.
-This is because our field is still relatively young (and thus high margin), but that is changing quickly.
-The trend you need to get behind is not AI, but quality.
-Whoever in our industry fully internalizes the lessons of Fred Brooks, Deming and Juran first will win.
-Tooling such as suggested here would greatly further those ends, and give whatever language that builds them great advantage over their alternatives.
+The software industry is one of the few places aside from SoEs\footnote{State-Owned Enterprises} where a 2/3 rate of failure \cite{standish} is remotely tolerable.
+This is because our field is still relatively young (and thus can afford such waste due to high profit margins) but that is changing quickly.
+Programming communities that do not improve their tooling to enable writing programs with less defects will over time be out-competed by those that do.
 

### 2024-06-23 sciperl/paper.tex
more stuff

@@
         title  = "Bezier's software testing techniques",
-        note   = "TODO: chapter/verse"
+        note   = "ibid"
     }
@@
         url  = {https://github.com/teodesian/Pod-Coverage-More/blob/master/lib/Pod/Coverage/More.pm}
+    }
+    @misc{mltype,
+        author = "Harry Mairson",
+        year   = 1989,
+        journal = "Proceedings of the ACM",
+        url     = {https://dl.acm.org/doi/10.1145/96709.96748}
+    }
+    @misc{clippert,
+        title  = "Lambda Expressions vs. Anonymous Methods, pt. 5",
+        url    = {https://learn.microsoft.com/en-us/archive/blogs/ericlippert/lambda-expressions-vs-anonymous-methods-part-five},
+        author = "Eric Clippert",
+        year   = 2007
     }
@@
 
-The difficulty here is the scope of such an undertaking.
+The difficulty here is twofold: the scope of such an undertaking and the computational difficulty to do so.
 A lookup table of every subroutine in Perl's CORE\footnote{\texttt{perldoc perlfunc}} namespace's arguments mapping to perl's internal datatypes \footnote{see \texttt{perldoc perlguts}} would need to be built.
@@
 I found most of this out the "hard way" \cite{gsb} long ago attempting to solve this particular problem (and failing).
+
+Nevertheless, this is a problem which others have attempted to solve and done so with some success.
+Haskell, ML and Swift all have such type inference systems with the caveat of exponential worst-case execution times\cite{clippert}.
+This is because the problem is in fact the 3SAT problem\footnote{also known as the boolean satisfiability problem} which is NP-hard\cite{mltype}.
+
+Like all other NP-hard problems, they can be overcome by solving a simpler subset of the problem (aka "cheating").
+A type inferencer can be lifted from one of these languages, and if we can devise a means to studiously avoid the perverse cases it would be of great utility.
 
@@
 Doing this will identify where the most high-impact test coverage improvements can be made in the perl ecosystem.
-
-Yet another place where I would love to have a proof of concept.
 
@@
 Thankfully there is a solution which uses XS to accomplish this: \texttt{Devel::Probe}\footnote{https://metacpan.org/pod/Devel::Probe}.
-
-TODO: show off POC doing this with tests...
 

### 2024-06-23 sciperl/paper.tex
finally done

@@
 in a rigorous fashion, and identify places where our ability to reason could be improved via tooling.
-Prototypes of said tools will also be demonstrated.
 
@@
 Nevertheless, this is a problem which others have attempted to solve and done so with some success.
-Haskell, ML and Swift all have such type inference systems with the caveat of exponential worst-case execution times\cite{clippert}.
+Haskell, ML and Swift all have such type inference systems with the caveat of exponential worst-case compile times\cite{clippert}.
 This is because the problem is in fact the 3SAT problem\footnote{also known as the boolean satisfiability problem} which is NP-hard\cite{mltype}.
@@
 This of course ignores the reality that poor choice of type (such as an "Any" type, or unsigned when you need signs, etc) immediately voids any such guarantees.
-Boundary testing also remains necessary.
 However the argument is not entirely without merit, and there are various runtime constraint systems such as \texttt{Type::Tiny} attesting to this\footnote{Along with other efforts, such as those promoted by the Perl Types committee: https://perlcommunity.org/types/}.
 
-Typescript\footnote{a ECMAScript (aka Javascript) code generator} is probably the best example of a dynamic language figuring out how to give you such compile-time guarantees.
+Bolting on strict types to a dynamic language is not a subject to approach lightly, as many features like late binding and reflection become far more difficult.
+We should learn from existing successful attempts to do so.
+
+Typescript\footnote{a ECMAScript (aka Javascript) code generator} is the most successful example of a dynamic language figuring out how to bolt-on static types.
 Supposing everything else it interacts with is also generated by Typescript\footnote{Rarely ever the case in Javascript projects}, a good degree of confidence can be had.
-It's also likely that much of the code therein, particularly the types themselves, can be repurposed towards doing the same with perl.
 
@@
 Such subroutines usually behave poorly in threaded contexts.},
-such failures tend to go unnoticed when inured in this mental model.
+failures-by-value tend to go unnoticed when inured in this mental model.
 This is why a great many Perl::Critic policies exist to catch precisely such problems caused by external factors largely outside of a subroutine's control.
@@
 
-The sources of these faults outside of a subroutine's control are generally syscalls or FFI.
+The sources of these faults outside of a subroutine's control are generally from syscalls or FFI\footnote{Foreign function interface}.
 As such we will need some means of discerning whether or not the usage in a test behaves well under these scenarios:
@@
 Thankfully there is a solution which uses XS to accomplish this: \texttt{Devel::Probe}\footnote{https://metacpan.org/pod/Devel::Probe}.
+There is also a more fleshed-out and generic framework called \texttt{Devel::Events}\footnote{https://metacpan.org/pod/Devel::Events}.
 
@@
 This is also another place where AI techniques could be of great utility.
-While they are necessarily awful at writing \textit{correct} code \cite{copilot}, they're pretty great at classifying code.
+While they are necessarily inept at writing \textit{correct} code \cite{copilot}, they're pretty great at classifying code.
 As such we could likely build these lists of "Modules that do largely the same things on CPAN" in a mostly automated fashion.

### 2024-06-23 sciperl/paper.tex
more

@@
 
-\subsection{100\% coverage, or: how many angels can dance upon the head of a pin?}
+\subsection{Code coverage all the way down the stack}
 
@@
 There is also a more fleshed-out and generic framework called \texttt{Devel::Events}\footnote{https://metacpan.org/pod/Devel::Events}.
+An automatic fault injection and coverage reporter via these means has yet to be built.
 
@@
 The software industry is one of the few places aside from SoEs\footnote{State-Owned Enterprises} where a 2/3 rate of failure \cite{standish} is remotely tolerable.
-This is because our field is still relatively young (and thus can afford such waste due to high profit margins) but that is changing quickly.
+This is because our field is still relatively young (and thus can afford such waste due to high profit margins).
+Such a state of affairs will not persist indefinitely.
 Programming communities that do not improve their tooling to enable writing programs with less defects will over time be out-competed by those that do.
